Anxiety, avoidance, and sequential evaluation

### 2019-06-30 manuscripts/manuscript/manuscript.tex
minor changes.

In the following, we will demonstrate through simulation how our simple model can account for anxious behavior in laboratory-based studies of sequential learning and decision making. We will also show that our model is consistent with clinical theory describing the transition from pathological anxiety to depression. Unless otherwise noted, state and action values under varying degrees of pessimism were solved for using the value iteration dynamic programming method\citep{SuttonBarto2018}. All simulations were implemented in the Python programming language and the code is publicly available at \url{https://github.com/ndawlab/seqanx}.

@@

% Fix subsection spacing?

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_choice.png}}%
  }
  \par \textbf{Figure 4:} (A) The free choice premium paradigm from \cite{Leotti2011} and \cite{Leotti2014} with equal chance of outcome $R \in [1, -1]$. Non-anxious participants exhibit a preference for the free choice option (blue) despite it conferring no benefit over the fixed choice option (grey). (B) Pessimistic agents show an attenuated free choice bias. (Parameters: $\gamma = 1.0$, Inverse temperature, $\beta$, increased from 1 to 10 over 100 episodes. Each dot represents one of 500 simulated agents.)
\end{figure}

@@

Indeed, our account formalizes a longstanding strand of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory\citep{bandura1977} and the triple vulnerabilities model\citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. Similarly, the learned helplessness theories of anxiety\cite{alloy1990} claims pathological anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model (though stated here in terms of self-efficacy) can accommodate either variant, and they are by no means mutually exclusive.

@@

None of the previous discussion constitutes a claim about the algorithmic and neural bases of pessimistic sequential evaluation. One possibility is that these symptoms reflect aberrations in model-based planning. Chronic worry is a symptom frequently observed in pathological anxiety, and could be construed as a deviant form of future action planning wherein the roll-out of future actions is biased towards bad- or worst-case contingencies. In line with the present results, higher levels of worry is correlated with reduced perceived control, increased doubt in self-efficacy in response to threat, and exaggerated threat appraisal\cite{Berenbaum2010}. In sum, anxious symptoms may in part result from biased planning where cost-minimizing action is discounted and

%% Need to clean the above up

@@

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our model of pessimistic sequential evaluation is, by definition, a model of prospective cognition. As such, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can neither account for the compulsive behaviors of obsessive-compulsive disorder nor the memory disturbances of post-traumatic stress disorder (though recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD\citep{gallagher2014a, gallagher2014b}). Importantly, other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. As such, the present results are transdiagnostic and not limited to one particular diagnosis.

>>>>

In the following, we will demonstrate through simulation how our simple model can account for anxious behavior in laboratory-based studies of sequential learning and decision making. We will also show that our model is consistent with clinical theory describing the transition from clinical anxiety to depression. Unless otherwise noted, state and action values under varying degrees of pessimism were solved for using the value iteration dynamic programming method\citep{SuttonBarto2018}. All simulations were implemented in the Python programming language and the code is publicly available at \url{https://github.com/ndawlab/seqanx}.

@@

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_choice.png}}%
  }
  \par \textbf{Figure 4:} (A) The free choice premium paradigm from \cite{Leotti2011} and \cite{Leotti2014} with equal chance of outcome $R \in [1, -1]$. Non-anxious participants exhibit a preference for the free choice option (blue) despite it conferring no benefit over the fixed choice option (grey). (B) Pessimistic agents show an attenuated free choice bias. (Parameters: Q-learning with $\gamma = 1.0$, and inverse temperature, $\beta$, increased from 1 to 10 over 100 episodes. Each dot represents one of 500 simulated agents.)
\end{figure}

@@

Indeed, our account formalizes a longstanding strand of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory\citep{bandura1977} and the triple vulnerabilities model\citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. Similarly, the learned helplessness theories of anxiety\cite{alloy1990} claims clinical anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model (though stated here in terms of self-efficacy) can accommodate either variant, and they are by no means mutually exclusive.

@@

% not sure if this paragraph is good

None of the previous discussion constitutes a claim about the algorithmic and neural bases of pessimistic sequential evaluation. One possibility is that these symptoms reflect aberrations in model-based planning. Model-based planning refers to the process of sequentially evaluation of trajectories of action, which may go awry in anxiety if successful avoidance actions are discounted or otherwise downweighted. Chronic worry is a symptom frequently observed in clinical anxiety, and bears resemblance to aberrant model-based planning wherein the trajectories are biased towards bad- or worst-case contingencies. In line with the present results, chronic worry is associated with reduced perceived control, diminished belief in self-efficacy in response to threat, and exaggerated threat appraisal\cite{Berenbaum2010}. This suggests that clinical anxiety may in part result from planning processes gone awry.

@@

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our model of pessimistic sequential evaluation is, by definition, a model of prospective cognition. As such, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can neither account for the compulsive behaviors of obsessive-compulsive disorder nor the memory disturbances of post-traumatic stress disorder (though recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD)\citep{gallagher2014a, gallagher2014b}. Importantly, other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. As such, the present results are transdiagnostic and not limited to one particular diagnosis.
